Wheels of Misfortune: Exploring Auto Theft Trends in Ottawa

### 2024-04-25 paper/paper.qmd
updated paper

---
title: "Wheels of Misfortune: Exploring Auto Theft Trends in Ottawa"
subtitle: "An Analysis on top car brands stolen, recovery rate, and top neighborhoods with high theft rates"
author: 
  - Catherine Punnoose
thanks: "Code and data are available at: https://github.com/catherinee1216/AutoTheft_Ottawa.git."
date: today
date-format: long
abstract: "The surge in auto theft rates across Canada has prompted car owners to become more vigilant regarding their vehicle's security. This paper examines auto theft occurrences within Canada's capital, Ottawa, Ontario. Focusing on the years 2018-2023, the paper considers the types of vehicles, brands, and neighborhoods targeted within Ottawa as well as the recovery rate. Additionally, the study aims to discuss reasons for auto theft and comparisons to data in province."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

library(tidyverse)
library(janitor)
library(kableExtra)
library(ggmap)
library(arrow)
library(knitr)
library(tinytex)
library(readr)
library(janitor)
library(dplyr)

@@

# Introduction

@@

This paper aims to provide a deep analysis of auto theft rates in Ottawa, Canada. The estimand for this paper is to quantify the impact of various factors on the occurrence and frequency of auto thefts in Ottawa, ultimately aiming to understand the dynamics driving the escalating trend in auto theft rates within the city. Subsequently, the @Data section focuses on the methodology for data collection and data cleaning. The @Results section delves into a comprehensive analysis of auto theft patterns in Ottawa from the years 2018-2022. Finally, in the @Discussion section, the paper explores potential drivers behind the escalation in auto theft rates and comparing the data to trends across Ontario.

@@

The data was cleaned by using the open source statistical programming language R [@r], with libraries `tidyverse` [@tidyverse], `ggplot2` [@ggplot2], `dplyr` [@dplyr], `readr` [@readr], `tibble` [@tibble], `here` [@here], `kableExtra` [@kableExtra], `janitor` [@janitor], and `knitr` [@knitr].

@@

# Results

@@

```{r}
#| label: fig-theft-recovery
#| fig-cap: "Recovery Rates from 2018-2023"
#| echo: false
#| warning: false
#| message: false

```{r}
#| label: fig-theft-neighborhoods
#| fig-cap: "Top 5 Neighborhoods with Most Auto Thefts from 2018-2023"
#| echo: false
#| warning: false
#| message: false

# Filter for top 10 neighborhoods with the highest auto theft rates from 2018-2023
theft_recovery <- auto_Data |>
  select(neighbourhood) |> 
  count(neighbourhood) |> 
  filter(n > 180)

theft_recovery |>
  ggplot(aes(x = neighbourhood, y = n, fill = n)) +  # Use n as the fill aesthetic
  geom_bar(stat = "identity") +
  theme_minimal() + theme(legend.position = "bottom") + 
  labs(x = "Vehicle Brand", y = "Count", fill = "Count")

@@

@fig-theftrecovery illustrates the number of cars recovered and unrecovered from 2018 to 2023. Overall, it is evident that more cars are recovered, except for 2023, where more cars were left unrecovered compared to those recovered. From 2018 to 2021, the number of cars recovered consistently surpassed those left unrecovered. Specifically, in 2018, 570 cars were recovered out of 914 (as shown in Figure 1), reflecting a 62% recovery rate. Similarly, in 2019, with a total of 1,012 cars stolen (as shown in Figure 1), 646 were recovered (as shown in Figure 3), maintaining a recovery rate of 63%.

In 2020, 529 out of 891 stolen cars were recovered, resulting in a 59% recovery rate. The following year, 2021, witnessed a slightly lower recovery rate of 53%, with 623 cars recovered out of 1,157 stolen. Notably, 2022 marked the highest number of both recovered and unrecovered cars, totaling 1,080 and 1,079 respectively. As per [FIGURE 1], a total of 2,159 cars were stolen in Ottawa in 2022. [FIGURE 3] indicates that 1,080 out of 2,159 were recovered, equating to a 50% recovery rate for 2022.

However, 2023 stands out as the only year where the number of cars recovered is fewer than those left unrecovered. In that year, a total of 693 cars were stolen (as shown in Figure 1), and only 334 cars were recovered (as shown in Figure 3), resulting in a recovery rate of 48%. [FIGURE 3] illustrates that between 2018 and 2023, the recovery rate averages roughly between 60% and 50%.

@@

# Discussion

@@

Although this paper focuses specifically on auto theft rates in Ottawa, how different are the trends compared to Ontario? According to a report by Equite, the majority of vehicle thefts in Ontario were vehicles made in 2017 or newer [@citeEquite]. This can be seen in [FIGURE] where newer vehicles made in 2017 or newer had a higher rate of theft in Ottawa.

@@

In conclusion, the surge in auto theft incidents, both in Ottawa and across Canada, underscores a pressing issue that demands attention. The escalation in theft rates poses a significant threat to public safety, prompting concerns about the effectiveness of current preventative measures and law enforcement efforts. The analysis conducted in this paper sheds light on the auto theft landscape specifically in Ottawa, offering valuable insights into the trends and patterns observed between 2018 and 2023. By examining variables such as the types of vehicles targeted, the recovery rates, and the geographical distribution of thefts across different neighborhoods, this research provides a comprehensive understanding of the dynamics driving the escalating trend in auto theft. Moving forward, addressing the root causes behind the surge in auto theft requires a multifaceted approach that combines effective law enforcement strategies, community engagement initiatives, and technological solutions to deter criminals and protect vulnerable vehicles. By leveraging the findings of this study and collaborating with stakeholders, policymakers, and law enforcement agencies can develop targeted interventions to curb the rising tide of auto theft and safeguard the well-being of Ottawa residents.

\appendix

# Appendix {-}

# Additional data details

>>>>

---
title: "Wheels of Misfortune: Exploring Auto Theft Trends in Ottawa"
subtitle: "An Analysis on Top Car Brands Stolen, Recovery Rate, and Top Neighborhoods with High Theft Rates"
author: 
  - Catherine Punnoose
thanks: "Code and data are available at: https://github.com/catherinee1216/AutoTheft_Ottawa.git."
date: today
date-format: long
abstract: "The surge in auto theft rates across Canada has prompted car owners to become more vigilant regarding their vehicle's security. This paper examines auto theft occurrences within Canada's capital, Ottawa, Ontario. Focusing on the years 2018-2023, the paper considers the types of vehicles, brands, and neighborhoods targeted within Ottawa as well as the recovery rate. Additionally, the study aims to discuss reasons for auto theft and comparisons to data in province."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

library(tidyverse)
library(janitor)
library(kableExtra)
library(arrow)
library(knitr)
library(tinytex)
library(readr)
library(janitor)
library(dplyr)
library(ggplot2)

@@

# Introduction {#sec-intro}

@@

This paper aims to provide a deep analysis of auto theft rates in Ottawa, Canada. The estimand for this paper is to quantify the impact of various factors on the occurrence and frequency of auto thefts in Ottawa, ultimately aiming to understand the dynamics driving the escalating trend in auto theft rates within the city. Subsequently, the @sec-data section focuses on the methodology for data collection and data cleaning. The @sec-Results section delves into a comprehensive analysis of auto theft patterns in Ottawa from the years 2018-2022. Finally, in the @sec-Discussion section, the paper explores potential drivers behind the escalation in auto theft rates and comparing the data to trends across Ontario.

@@

The dataset used for this paper was provided by Ottawa Police [@citeOttawa]. The dataset contains theft of motor vehicle occurrencesfrom the years 2018-2023.The data was cleaned by using the open source statistical programming language R [@citeR], with libraries `tidyverse` [@tidyverse], `ggplot2` [@ggplot2], `dplyr` [@dplyr], `readr` [@readr], `tibble` [@tibble], `kableExtra` [@kableExtra], `janitor` [@janitor], arrow [@arrow] ,and `knitr` [@knitr].

@@

# Results {#sec-Results}

@@

@@

@fig-theftrecovery illustrates the number of cars recovered and unrecovered from 2018 to 2023. Overall, it is evident that more cars are recovered, except for 2023, where more cars were left unrecovered compared to those recovered. From 2018 to 2021, the number of cars recovered consistently surpassed those left unrecovered. Specifically, in 2018, 570 cars were recovered out of 914 (@fig-averageTheftYear), reflecting a 62% recovery rate. Similarly, in 2019, with a total of 1,012 cars stolen (@fig-averageTheftYear), 646 were recovered (@fig-theftrecovery), maintaining a recovery rate of 63%.

In 2020, 529 out of 891 stolen cars were recovered, resulting in a 59% recovery rate. The following year, 2021, witnessed a slightly lower recovery rate of 53%, with 623 cars recovered out of 1,157 stolen. Notably, 2022 marked the highest number of both recovered and unrecovered cars, totaling 1,080 and 1,079 respectively. As per @fig-averageTheftYear, a total of 2,159 cars were stolen in Ottawa in 2022. @fig-theftrecovery indicates that 1,080 out of 2,159 were recovered, equating to a 50% recovery rate for 2022.

However, 2023 stands out as the only year where the number of cars recovered is fewer than those left unrecoverable. A total of 693 cars were stolen (as shown in @fig-averageTheftYear), and only 334 cars were recovered (as shown in @fig-theftrecovery), resulting in a recovery rate of 48%. @fig-theftrecovery illustrates that between 2018 and 2023, the recovery rate averages roughly between 60% and 50%.

@@

# Discussion {#sec-Discussion}

@@

Although this paper focuses specifically on auto theft rates in Ottawa, how different are the trends compared to Ontario? According to a report by Equite, the majority of vehicle thefts in Ontario were vehicles made in 2017 or newer [@citeEquite]. This can be seen in @fig-vehicleYearStolen where newer vehicles made in 2017 or newer had a higher rate of theft in Ottawa.

@@

In conclusion, the surge in auto theft incidents, both in Ottawa and across Canada, is a pressing issue that demands attention. The escalation in theft rates poses a significant threat to public safety, prompting concerns about the effectiveness of current preventative measures and law enforcement efforts. The analysis conducted in this paper sheds light on the auto theft landscape specifically in Ottawa, offering insights into the trends and patterns observed between 2018 and 2023. By examining variables such as the types of vehicles targeted, the recovery rates, and the geographical distribution of thefts across different neighborhoods, this research provides a comprehensive understanding of the dynamics driving the escalating trend in auto theft. Moving forward, addressing the root causes behind the surge in auto theft requires a multifaceted approach that combines effective law enforcement strategies, community engagement initiatives, and technological solutions to deter criminals and protect vulnerable vehicles. By leveraging the findings of this study and collaborating with stakeholders, policymakers, and law enforcement agencies can develop targeted interventions to curb the rising tide of auto theft and safeguard the well-being of Ottawa residents.
